---
title: "Practical Machine Learning Course Project"
author: "Anton Hesse"
date: "6/21/2020"
output: html_document
---

Practical Machine Learning Course Project

GitHub Repo: https://github.com/ahesse2567/practicalmachinelearning


## Background Information
Physical activity monitors such as heart rate monitors have been commonplace for a long while. However, newer devices may have the capability to analyze weightlifting exercises to determine the quality of each repetition. That is, was the person performing the weightlifting exercise doing it incorrectly because of improper technique knowledge? Or, are they succumbing to fatigue and making improper compensations? This may be useful for both teaching exercises, assessing and evaluating training, and predicting injury risk.

In this particular dataset, particpants were performing bicep curls in the correct and in several incorrect manners. Full information on the study and the dataset can be found here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har#related_publications. Additional information is found below in the assignment description.

## Assignment description

"What you should submit:

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases."


## Reproducibility
The following is the list of packages and the seed so others can verify the results

```{r warning=FALSE, error=FALSE}
# load packages and set the seed
library(tidyverse); library(caret); library(magrittr); library(car); library(corrplot)
library(randomForest); library(rattle);
set.seed(1234)
```


## Setting up the data
First, we need to import the data from the URL. Next we will split the training data into training and testing datasets. The "testing" url will serve as the validation set.
```{r warning=FALSE, error=FALSE}
# create training and testing sets
training <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
inTrain <- createDataPartition(training$classe, p = 0.7, list = FALSE)
training <- training[inTrain,]
testing <- training[-inTrain,]

dim(training)
dim(testing)

# create validation set
validation <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
dim(validation)
```

## Exploratory Data Analysis and Data Cleaning

``` {r warning=FALSE, error=FALSE}
summary(training)
```
We have to deal with NA's, blanks, and #DIV0!'s. We will identify how many NA values there are first.

``` {r warning=FALSE, error=FALSE}
(naList <- colnames(training)[colSums(is.na(training)) > 0])
(noNAList <- colnames(training)[colSums(is.na(training)) == 0])

a <- sapply(training, function(x) sum(is.na(x))/nrow(training))
b <- which(a > 0)
a[b]
```
98% of values in columns w/ NA values are NA values. This is too many to impute and we would therefore exclude those columns.

``` {r warning=FALSE, error=FALSE}
training <- training[, apply(training, 2, function(x) !any(is.na(x)))]
testing <- testing[, apply(testing, 2, function(x) !any(is.na(x)))]
validation <- validation[, apply(validation, 2, function(x) !any(is.na(x)))]
dim(training)
dim(testing)
dim(validation)
```

There are also some predictors that appear to be for bookkeeping purposes only. We will identify and remove theose

``` {r warning=FALSE, error=FALSE}
colnames(training)
training <- training[,-c(1:5)] # remove variables that are likely for bookkeeping only
str(training) # several variables need to be converted from factors into numeric
```

From looking at the structure of the data it's clear that several variables are factors when they should be numeric.

``` {r warning=FALSE, error=FALSE}
training[,-length(training)] %<>% # remove all but last variable since it's the classe
  mutate_if(is.factor,as.numeric)
str(training)
```

Changing the data from factor to numeric appeared to also resolve the NA's and #DIV0!'s. Repeat for testing and validation sets

``` {r warning=FALSE, error=FALSE}
testing <- testing[,-c(1:5)] # remove variables that are likely for bookkeeping only
testing[,-length(testing)] %<>% # remove all but last variable since it's the classe
  mutate_if(is.factor,as.numeric)

# repeat for validation
validation <- validation[,-c(1:5)] # remove variables that are likely for bookkeeping only
validation[,-length(validation)] %<>% 
  mutate_if(is.factor,as.numeric)
```

Now we will examine how many variables have near zero variation because they likely wouldn't contribute much information to the final model.

``` {r warning=FALSE, error=FALSE}
names(training)[nearZeroVar(training)]
nzv <- nearZeroVar(training)
var(training[,nzv])
length(nzv) # 34 variables with nearZeroVar

training <- training[,-nzv] # remove variables with near zero variation
dim(training)

# repeat for testing and validation sets
testing <- testing[,-nearZeroVar(testing)]
validation <- validation[,-nearZeroVar(validation)]
```

## Data Visualization
Now that the data is cleaned, we will correlate predictors to assess covariation
``` {r warning=FALSE, error=FALSE}
corObj <- cor(testing[-length(testing)])

corrplot(corObj, type = "lower", order="hclust",
         tl.col="black", tl.srt = 45)
```
Covariation exists within this dataset, but it appears to be minimal.

## Model Creation

We will test four models:
1. Classification and Regression Trees (CART)
2. Random Forests
3. Bagging
4. Boosting

1. Classification and Regression Trees (CART):

``` {r warning=FALSE, error=FALSE}
modCART <- train(classe ~ ., data = training,
                 preProcess = c("center", "scale"), method = "rpart")
modCART
```
The accuracy with the training set was only 53%.

2. Random Forests:
``` {r warning=FALSE, error=FALSE}
train_Control <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
modRF <- train(classe ~., data = training, preProcess = c("center", "scale"),
               method="rf", trControl = train_Control)
modRF
```
Very high accuracy with random forests at 99.1%

3. Bagging:
``` {r warning=FALSE, error=FALSE}
modBag <- train(classe ~., data = training, preProcess = c("center", "scale"),
                method = "treebag", trControl = train_Control)
modBag
```
Very high accuracy, but not quite as high as random forests

4. Boosting:
``` {r warning=FALSE, error=FALSE}
modBoosting <- train(classe ~ ., data=training, method = "gbm",
                     trControl = train_Control, verbose = FALSE)
modBoosting
```
Once again, very high accuracy, but not quite as high as random forests

## Estimating model accuracy with testing datasets and final prediction

``` {r warning=FALSE, error=FALSE}
predict_CART <- predict(modCART, testing)
predict_RF <- predict(modRF, testing)
predict_Bag <- predict(modBag, testing)
predict_Boosting <- predict(modBoosting, testing)

(cm_CART <- confusionMatrix(testing$classe, predict_CART)) # only 50% correct
(cm_RF <- confusionMatrix(testing$classe, predict_RF)) # perfect prediction
(cm_Bag <- confusionMatrix(testing$classe, predict_Bag)) # perfect prediction
(cm_Boosting <- confusionMatrix(testing$classe, predict_Boosting)) # 99.1% correct
```

Both random forests and bagging gave perfect predictions. We will use the random forest model on the validation set. Perfect accuracy might be due to overfitting. Regardless, the out- of-sample error rate is expected to be 100-100 = 0.

``` {r warning=FALSE, error=FALSE}
predict_validation <- predict(modRF, validation)
predict_validation
```